Tackling inconsistencies among freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2022-03-26 Paper/Submission_new/Main_Document-Latex.tex
26.03: Updated paper files, most recent in Submission_new

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{etoolbox}
\usepackage{hhline}
\usepackage{lmodern}
\usepackage[
backend = biber,
natbib,
citestyle = authoryear,
bibstyle = apa,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

@@

% raplace "and" in authoryear in-text citations with "&" 
\renewcommand*{\finalnamedelim}{%
  \ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}%
  \addspace\&\space}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution}
\author[1]{Stefan Kunz}
\author[2]{Ben J. Kefford}
\author[3]{Astrid Schmidt-Kloiber}
\author[4]{Christoph D. Matthaei}
\author[5]{Philippe Usseglio-Polatera}
\author[3]{Wolfram Graf}
\author[6]{N. LeRoy Poff}
\author[7]{Leon Metzeling}
\author[8]{Laura Twardochleb}
\author[9]{Charles P. Hawkins}
\author[1]{Ralf B. Schäfer}
\affil[1]{Institute for Environmental Sciences, University of Koblenz-Landau, Landau, Germany}
\affil[2]{Centre for Applied Water Science, Institute for Applied Ecology, University of Canberra, Canberra, Australia}
\affil[3]{Institute of Hydrobiology and Aquatic Ecosystem Management, University of Natural Resources and Life Sciences Vienna (BOKU), Vienna, Austria}
\affil[4]{Department of Zoology, University of Otago, Dunedin, New Zealand}
\affil[5]{University of Lorraine, CNRS, LIEC, Metz, France}
\affil[6]{Department of Biology, Colorado State University, Fort Collins, USA}
\affil[7]{Environment Protection Authority Victoria, Applied Sciences Division, Macleod, Australia}
\affil[8]{Department of Fisheries and Wildlife, Michigan State University, East Lansing, USA}
\affil[9]{Department of Watershed Sciences, National Aquatic Monitoring Center, and the Ecology Center, Utah State University, Logan, USA}
\date{}

@@

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere.  In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, standardised definitions for freshwater invertebrate traits are lacking, which impedes comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions among databases might influence identification of trait-environment relationships. In addition, taxonomic resolution between taxonomic composition data and trait data can differ, making aggregation of traits necessary. However, at present, it is unknown how different trait aggregation approaches compare with expert-assigned trait affinities.
% In this study we aimed to identify discrepancies in trait definitions across freshwater invertebrate databases and to compare trait aggregation methods to identify how phylogenetic structure and trait variability influence results from different aggregation methods.

2. We describe discrepancies in the definitions of traits used to create freshwater invertebrate trait databases in Europe, North America, New Zealand, and Australia. Based on our comparisons of these trait databases, we established four novel trait datasets by harmonising trait definitions of commonly used traits. Next, we used two of these datasets to compare aggregated traits obtained by different aggregation methods with traits assigned by experts, both at the family level. The trait aggregation methods we compared used either the mean or the median and different weightings. We further explored the effects of harmonisation and trait aggregation by re-analysing data from a case study.

3. We found that among trait databases, trait definitions often differed because varying numbers of traits were used to describe groups of related traits (e.g., respiration traits) and the focus of descriptions for groups of related traits also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially when aggregation was based on the median.
% ,and that trait coding method, uneven phylogenetic structure, and high trait variability can influence aggregated trait affinities. 
We further showed that harmonised and aggregated data performed similarly to non-aggregated and harmonised data from the case study in identifying trait–environment relationships, but were only partially able to yield similar values for functional diversity metrics in the case study.

4. By identifying discrepancies in trait definitions we hope to motivate the development of standardised definitions for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\keywords{trait definitions, trait aggregation, data synthesis, trait-environment relationships, large-scale comparisons}

@@

Explaining and predicting how communities are shaped by environmental factors is a primary goal of ecology. Species traits are measurable properties of an organism (\cite{mcgill_rebuilding_2006}), and comparing communities based on the traits possessed by species may facilitate testing of a range of ecological hypotheses (\cite{statzner_perspectives_2001}). Traits are assumed to be adaptations (e.g., physiological, behavioural) of organisms to their environment, which represent either direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}). In addition to providing local-scale mechanistic-based expressions of species-environment relationships, trait-based approaches might be suitable for large-scale analyses of environmental risks to communities because the occurrence and distribution of traits are less constrained by biogeographic boundaries than that of taxa (\cite{baird_toward_2011, bonada_taxonomic_2007}).

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, contiTraitbasedApproachAssess2014, poff_developing_2010, szocs_effects_2014}), for predicting changes in ecosystem functioning and trophic dynamics (\cite{gutierrez-canovasPopulationsHighvaluePredators2021, vos_taxonomic_2017}), for developing indices or diagnostic tools for biomonitoring (\cite{beketov_spear_2009, doledecInvertebrateTraitsBiomonitoring2008a, mondyUsingConditionalTree2013}), but see also \citet{hamiltonLimitationsTraitbasedApproaches2020} for a critical perspective. In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{gayraudInvertebrateTraitsBiomonitoring2003, keffordIntegratedDatabaseStream2020a, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015,
statznerConservationTaxonomicBiological2007, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. In fact, the first studies on the link between habitat utilization and traits in the Upper Rhône river initiated the compilation of aquatic invertebrate trait information (\cite{bournaudApproachSynthesisQualitative1992,usseglio-polateraTheoreticalHabitatTemplets1994}). At large scales, consistent responses of traits to environmental stressors have been reported, e.g. traits like small body size, multiple life cycles per year, or swimming have been related to resilience and resistance to flow disturbances or retreating glacier cover (\cite{bonada_taxonomic_2007, brown_functional_2018}). To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterised trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesise invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

@@

Finally, discrepancies in the taxonomic resolutions (e.g., species, genus, or family) used when comparing among trait databases, or when linking observed taxonomic data to trait databases, presents another challenge (\cite{gayraudInvertebrateTraitsBiomonitoring2003, doledecAccurateDescriptionAbundance2000}). When taxonomic composition or data in a trait database are at a more precise taxonomic level than other trait data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). If trait information is only available at more precise taxonomic levels than for taxonomic composition, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing.

These challenges motivated us to assess the characteristics and comparability of different freshwater invertebrate trait databases with the goal of facilitating future syntheses. We aimed to identify methods of improving the comparability of trait data across databases, despite differences in trait definitions and the taxonomic level of data in three ways: (1) by describing discrepancies in trait definitions between trait databases from Europe, North America, Australia, and New Zealand; (2) by harmonising trait definitions across databases to provide consistent grouping features for comparison across geographic locations; and (3) by comparing the influence of different trait aggregation methods and grouping-feature harmonisation on the detection of trait–environment relationships.

@@

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. Discrepancies in trait definitions served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features present in the existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation methods on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of phylogenetic structure and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected grouping features that are commonly applied in trait-based ecological studies and describe different aspects of the biology of a species: life history (Voltinism), morphology (Body form, Size), physiology (Feeding mode, Respiration), mobility (Locomotion), and reproduction (Oviposition). A criterion for trait selection was that the traits occurred in all databases, to allow for an analysis of potential differences in trait definitions and of the influence of grouping-feature harmonisation on trait-environment relationships. Thus, we only used biological traits and omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database.

@@

All data processing and RDA analyses were carried out using R version 3.6.1 (\cite{cite_R}). RDA was computed using the vegan package (\cite{cite_vegan}). Raw data and the R code for data processing and grouping feature harmonisation are located in the Github repository: \url{https://github.com/KunzstLD/Invertebrate_traits}. Scripts and data to reproduce the trait aggregation, comparison of aggregated and assigned traits, and the RDA analysis are located in the Github repository: \url{https://github.com/KunzstLD/Trait-aggregation}.

@@

\subsubsection*{Discrepancies in invertebrate trait definitions across databases}

@@

The definitions of grouping features also varied across databases in their differentiation into traits and in their focus, both which can lead to discrepancies in trait definitions (see Table \ref{tab:trait_definitions} and Table \ref{stab:trait_definitions} for a summary of discrepancies in trait definitions). Varying levels of differentiation among the trait databases were present in all investigated grouping features (Table \ref{tab:trait_definitions}, Table \ref{tab:trait_databases_coding_differentiation}, and Table \ref{stab:trait_definitions}).

For example, for the grouping feature feeding mode, discrepancies arose because traits were assigned in different ways. The Tachet database defined predators as carvers, engulfers, and swallowers. By contrast, the CONUS database defined predators as engulfers and carnivorous piercers. In turn, the Tachet database defined piercers as a separate trait encompassing both herbivorous and carnivorous piercers. Furthermore, the feeding mode in the freshwaterecology.info database was primarily described in terms of type of food (except for filterers), whereas the other databases described feeding mode in terms of strategies of food acquisition. For example, the freshwaterecology.info database defined predator as "eating from prey", and the other databases used the mouthpart morphology as the basis for their definitions. The Tachet database captured the food source in a separate grouping feature. Locomotion definitions also differed in focus among databases. The New Zealand and freshwaterecology.info databases described locomotion as an organism’s way of movement, the Tachet database as substrate relation and locomotion, the Vieira database as how organisms deal with flow, the Australia database as attachment, and the CONUS database as not only the method of movement but also the location of movement. Similarly, databases differed in the way they characterise reproduction traits. In the freshwaterecology.info and Tachet databases, reproduction was captured in one grouping feature and was defined as location of oviposit clutches and mode of reproduction. The Vieira database provided information on the oviposition location but not on reproductive behaviour. The Australian database reported grouping features for reproductive behaviour but also on oviposition location separately. The New Zealand database distinguished three grouping features related to reproduction: reproductive technique, oviposition location (e.g. water surface, terrestrial), and egg/egg mass (e.g. free, cemented).

@@

The percentage of differing cases of aggregated versus assigned trait affinities varied by method between 21.9 and 27.6 \% for the Australian dataset and between 15.5 and 44.8 \% for the North American dataset (Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). In general, median aggregation methods yielded fewer cases with differences compared with mean aggregation methods. However, median aggregation methods produced, on average, greater absolute differences for both datasets. Standard deviations of absolute differences were similar for all tested aggregation methods. Maximum differences of 1 (i.e., the maximum possible difference) occurred for all investigated grouping features in both datasets (Figure \ref{fig:diff_aggr_traits_combined}).

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=12cm]{Deviances_trait_agg_combined.png}
  \caption{Cases (factor combination of investigated families and traits) where differences occurred between aggregated traits and expert assigned traits at family level for the Australian and North American dataset. Violin plots - mirrored density plots - show the density of the absolute trait affinity differences for the grouping features feeding mode and body size. Results for the North American dataset for the grouping features locomotion, respiration, and voltinism are shown in the supporting information (Figure \ref{fig:diff_aggr_traits_pyne}). Absolute differences in trait affinities are depicted as grey dots. \textit{n} denotes the number of cases per comparison where differences occurred. The blue dot indicates the mean of absolute differences, and the error bars show the standard deviation. The grey vertical lines show the 25\textsuperscript{th}, 50\textsuperscript{th} and 75\textsuperscript{th} quantiles of the density estimate. Abbreviations: AUS, Australia; NA, North America.}
  \label{fig:diff_aggr_traits_combined}
\end{figure}

@@

RDA using harmonised, but not aggregated, grouping features identified some, but not all, of the same traits that were identified by the original analysis (Figures \ref{fig:violin_plot_species_sc} and \ref{fig:boxplot_scores_on_constrained_axis}). The original RDA in \citet{szocs_effects_2014} indicated that downstream sites (higher salinity) were characterised by the traits: ovoviviparity, multivoltinism, long life-cycle duration ($> 1$ year), shredder, and gill respiration, whereas RDA with the harmonised data identified the first three traits but did not include shredder or gill respiration traits. The original RDA indicated that upstream sites (lower salinity) were characterised by the traits: univoltinism, oviposition in clutches, and short life-cycle duration ($< 1$ year), and RDA with the harmonised data characterised upstream sites by univoltine taxa with short life-cycle durations that lay their eggs in an aquatic environment (aquatic eggs). This difference in the oviposition trait is not substantial because the trait aquatic eggs of the harmonised grouping feature oviposition was derived by combining the trait oviposition in clutches and other related traits (Table \ref{tab:traits_harmonisation}).

RDA based on harmonised traits that were aggregated to family level also produced results similar to the original analysis but with fewer traits distinguishing upstream and downstream sites (Figures \ref{fig:violin_plot_species_sc} and \ref{fig:boxplot_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} characterised the downstream sites with the same traits as the original analysis except the trait shredder. As for the RDA with harmonised but not aggregated data, RDA with these three aggregation methods characterised upstream sites by the traits univoltinism, short life-cycle duration, and aquatic eggs. The \textit{stepwise\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}} methods produced the same results except that none of the life-cycle traits characterised upstream or downstream sites. Thus, compared with the original study’s findings, aggregating by the direct and weighted methods yielded less change than the stepwise methods in interpretation of the RDA results. Given the criticism of approaches relying on CWM traits we also applied fourth corner analysis with the original, harmonised, and aggregated trait datasets to detect relationships between traits and conductivity (methods and results in the supporting information \ref{sec:effects_harmonisation_trait_env_rel}).

>>>>

\documentclass[12pt]{article} 
%\usepackage[a4paper, margin = 1n]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{etoolbox}
\usepackage{hhline}
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage[
backend = biber,
natbib,
citestyle = authoryear,
bibstyle = apa,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

@@

% replace "and" in authoryear in-text citations with "&" 
\renewcommand*{\finalnamedelim}{%
  \ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}%
  \addspace\&\space}

@@

% custom style for first page
\fancypagestyle{firstpage}
{
    \fancyhead[L]{Data-related issues in trait databases}    
    \fancyhead[R]{}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\thispagestyle{firstpage}
\title{Tackling inconsistencies among freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution}
\author[1]{Stefan Kunz}
\author[2]{Ben J. Kefford}
\author[3]{Astrid Schmidt-Kloiber}
\author[4]{Christoph D. Matthaei}
\author[5]{Philippe Usseglio-Polatera}
\author[3]{Wolfram Graf}
\author[6]{N. LeRoy Poff}
\author[7]{Leon Metzeling}
\author[8]{Laura Twardochleb}
\author[9]{Charles P. Hawkins}
\author[1]{Ralf B. Schäfer}
\affil[1]{Institute for Environmental Sciences, University of Koblenz-Landau, Landau, Germany}
\affil[2]{Centre for Applied Water Science, Institute for Applied Ecology, University of Canberra, Canberra, Australia}
\affil[3]{Institute of Hydrobiology and Aquatic Ecosystem Management, University of Natural Resources and Life Sciences Vienna (BOKU), Vienna, Austria}
\affil[4]{Department of Zoology, University of Otago, Dunedin, New Zealand}
\affil[5]{University of Lorraine, CNRS, LIEC, Metz, France}
\affil[6]{Department of Biology, Colorado State University, Fort Collins, USA}
\affil[7]{Environment Protection Authority Victoria, Applied Sciences Division, Macleod, Australia}
\affil[8]{Department of Fisheries and Wildlife, Michigan State University, East Lansing, USA}
\affil[9]{Department of Watershed Sciences, National Aquatic Monitoring Center, and the Ecology Center, Utah State University, Logan, USA}
\date{}

@@

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits potentially occur everywhere. In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports large scale studies. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. It is uncertain how harmonising varying trait definitions among databases might influence subsequent identification of trait-environment relationships. Furthermore, there have been few comparisons of trait aggregation approaches with expert-assigned trait affinities.

2. We describe inconsistencies in the definitions of traits used to create freshwater invertebrate trait databases in Europe, North America, New Zealand, and Australia. Based on our comparisons of these databases, we established four novel trait datasets by harmonising definitions of commonly used traits. Next, we used two of these datasets to compare aggregated traits obtained by different aggregation methods with traits assigned by experts, both at the family level. The trait aggregation methods we compared used either the mean or the median and different weightings. We further explored the effects of harmonisation and trait aggregation by re-analysing data from a case study.

3. We found that among databases, trait definitions often differed because varying numbers of traits were used to describe particular functions (e.g., respiration traits) and the way those functions were described also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases.

4. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially when traits were aggregated based on the median of trait values of taxa within a family.
% ,and that trait coding method, uneven phylogenetic structure, and high trait variability can influence aggregated trait affinities. 
% We further showed that harmonised and aggregated data performed similarly to non-aggregated and harmonised data from the case study in identifying trait–environment relationships, but were only partially able to yield similar values for functional diversity metrics in the case study.
The case study showed that harmonised and aggregated data identified similar trait-environment relationships to non-aggregated data. However, harmonised and aggregated data yielded only partially similar values for functional diversity metrics when compared to the case study results.

5. By identifying inconsistencies in trait definitions we hope to motivate the development of standardised definitions for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\keywords{trait definitions, trait aggregation, data synthesis, trait-environment relationships, large-scale comparisons}

@@

Explaining and predicting how communities are shaped by environmental factors is a primary goal of ecology. Species traits are measurable properties of an organism that strongly influence individual performance (\cite{mcgill_rebuilding_2006}), and comparing communities based on the traits possessed by species may facilitate testing of a range of ecological hypotheses (\cite{statzner_perspectives_2001}). Traits are assumed to be adaptations (e.g., physiological, behavioural) of organisms to their environment, which represent either direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}). In addition to providing local-scale mechanistic-based expressions of species-environment relationships, trait-based approaches might be suitable for large-scale analyses of environmental risks to communities because the occurrence and distribution of traits are less constrained by biogeographic boundaries than that of taxa (\cite{baird_toward_2011, bonada_taxonomic_2007}).

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, contiTraitbasedApproachAssess2014, poff_developing_2010, szocs_effects_2014}), for predicting changes in ecosystem functioning and trophic dynamics (\cite{gutierrez-canovasPopulationsHighvaluePredators2021, vos_taxonomic_2017}), for developing indices or diagnostic tools for biomonitoring (\cite{beketov_spear_2009, doledecInvertebrateTraitsBiomonitoring2008a, mondyUsingConditionalTree2013}), but see also \citet{hamiltonLimitationsTraitbasedApproaches2020} for a critical perspective. In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{gayraudInvertebrateTraitsBiomonitoring2003, keffordIntegratedDatabaseStream2020a, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015,
statznerConservationTaxonomicBiological2007, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. In fact, the first studies on the link between habitat use and traits in the Upper Rhône river initiated the compilation of aquatic invertebrate trait information (\cite{bournaudApproachSynthesisQualitative1992,usseglio-polateraTheoreticalHabitatTemplets1994}). At large scales, consistent responses of traits to environmental stressors have been reported, e.g. traits like small body size, multiple life cycles per year, or swimming have been related to resilience and resistance to flow disturbances or retreating glacier cover (\cite{bonada_taxonomic_2007, brown_functional_2018}). To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterised trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesise invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

@@

Finally, differences in the taxonomic resolutions (e.g., species, genus, or family) used when comparing among trait databases, or when linking observed taxonomic data to trait databases, presents another challenge (\cite{gayraudInvertebrateTraitsBiomonitoring2003, doledecAccurateDescriptionAbundance2000}). When taxonomic composition or data in a trait database are at a more precise taxonomic level than other trait data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). If trait information is only available at more precise taxonomic levels than for taxonomic composition, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing.

These challenges motivated us to assess the characteristics and comparability of different freshwater invertebrate trait databases with the goal of facilitating future syntheses. We aimed to identify methods of improving the comparability of trait data across databases, despite differences in trait definitions and the taxonomic level of data in three ways: (1) by describing inconsistencies in trait definitions between trait databases from Europe, North America, Australia, and New Zealand; (2) by harmonising trait definitions across databases to provide consistent grouping features for comparison across geographic locations; and (3) by comparing the influence of different trait aggregation methods and grouping-feature harmonisation on the detection of trait–environment relationships.

@@

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. Inconsistencies in trait definitions served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features present in the existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation methods on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of phylogenetic structure and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected grouping features that are commonly applied in trait-based ecological studies and describe different aspects of the biology of a species: life history (Voltinism), morphology (Body form, Size), physiology (Feeding mode, Respiration), mobility (Locomotion), and reproduction (Oviposition). A criterion for trait selection was that the traits occurred in all databases, to allow for an analysis of potential differences in trait definitions and of the influence of grouping-feature harmonisation on trait-environment relationships. Thus, we only used biological traits, i.e. traits describing survival ability in an environment (\cite{violleLetConceptTrait2007}), and omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database.

@@

All data processing, RDA analyses, and FD analysis were carried out using R version 3.6.1 (\cite{cite_R}). RDA was computed using the vegan package (\cite{cite_vegan}) and FD metrics were calculated with the FD package (\cite{cite_FD}).

@@

\subsubsection*{Inconsistencies in invertebrate trait definitions across databases}

@@

The definitions of grouping features also varied across databases in their differentiation into traits and in their focus, both which can lead to inconsistencies in trait definitions (see Table \ref{tab:trait_definitions} and Table \ref{stab:trait_definitions} for a summary of inconsistencies in trait definitions). Varying levels of differentiation among the trait databases were present in all investigated grouping features (Table \ref{tab:trait_definitions}, Table \ref{tab:trait_databases_coding_differentiation}, and Table \ref{stab:trait_definitions}).

For example, for the grouping feature feeding mode, inconsistencies arose because traits were assigned in different ways. The Tachet database defined predators as carvers, engulfers, and swallowers. By contrast, the CONUS database defined predators as engulfers and carnivorous piercers. In turn, the Tachet database defined piercers as a separate trait encompassing both herbivorous and carnivorous piercers. Furthermore, the feeding mode in the freshwaterecology.info database was primarily described in terms of type of food (except for filterers), whereas the other databases described feeding mode in terms of strategies of food acquisition. For example, the freshwaterecology.info database defined predator as "eating from prey", and the other databases used the mouthpart morphology as the basis for their definitions. The Tachet database captured the food source in a separate grouping feature. Locomotion definitions also differed in focus among databases. The New Zealand and freshwaterecology.info databases described locomotion as an organism’s way of movement, the Tachet database as substrate relation and locomotion, the Vieira database as how organisms deal with flow, the Australia database as attachment, and the CONUS database as not only the method of movement but also the location of movement. Similarly, databases differed in the way they characterise reproduction traits. In the freshwaterecology.info and Tachet databases, reproduction was captured in one grouping feature and was defined as location of oviposit clutches and mode of reproduction. The Vieira database provided information on the oviposition location but not on reproductive behaviour. The Australian database reported grouping features for reproductive behaviour but also on oviposition location separately. The New Zealand database distinguished three grouping features related to reproduction: reproductive technique, oviposition location (e.g. water surface, terrestrial), and egg/egg mass (e.g. free, cemented).

@@

The percentage of differing cases of aggregated versus assigned trait affinities varied by method between 16.7 and 23.4 \% for the Australian dataset and between 15.6 and 44.9 \% for the North American dataset (Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). In general, median aggregation methods yielded fewer cases with differences compared with mean aggregation methods. However, median aggregation methods produced, on average, greater absolute differences for both datasets. Standard deviations of absolute differences were similar for all tested aggregation methods. Maximum differences of 1 (i.e., the maximum possible difference) occurred for all investigated grouping features in both datasets (Figure \ref{fig:diff_aggr_traits_combined}).

@@

RDA using harmonised, but not aggregated, grouping features identified some, but not all, of the same traits that were identified by the original analysis (Figures \ref{fig:violin_plot_species_sc} and \ref{fig:boxplot_scores_on_constrained_axis}). The original RDA in \citet{szocs_effects_2014} indicated that downstream sites (higher salinity) were characterised by the traits: ovoviviparity, multivoltinism, shredder, long life-cycle duration ($> 1$ year), and gill respiration, whereas RDA with the harmonised data identified the first three traits but did not include long life-cycle duration ($> 1$ year) or gill respiration traits. The original RDA indicated that upstream sites (lower salinity) were characterised by the traits: univoltinism, oviposition in clutches, and short life-cycle duration ($< 1$ year), and RDA with the harmonised data characterised upstream sites by univoltine taxa that lay their eggs in an aquatic environment (aquatic eggs). This difference in the oviposition trait is not substantial because the trait aquatic eggs of the harmonised grouping feature oviposition was derived by combining the trait oviposition in clutches and other related traits (Table \ref{tab:traits_harmonisation}).

RDA based on harmonised traits that were aggregated to family level also produced results similar to the original analysis but with fewer traits distinguishing upstream and downstream sites (Figures \ref{fig:violin_plot_species_sc} and \ref{fig:boxplot_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg} characterised the downstream sites with the same traits as the original analysis except the trait life cycle duration $> 1$ year. As for the RDA with harmonised but not aggregated data, RDA with the five aggregation methods characterised upstream sites by the traits univoltinism and aquatic eggs. The \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}} methods produced the same results as the harmonised but not aggregated data. Thus, compared with the original study’s findings, aggregating by the mean and weighted methods yielded less change than the aggregation with the median in interpretation of the RDA results. Given the criticism of approaches relying on CWM traits we also applied fourth corner analysis with the original, harmonised, and aggregated trait datasets to detect relationships between traits and conductivity (methods and results in the Supplementary Materials).
